---
title: "Enrichment Analysis"
author: "Jasmine Chong, Jeff Xia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Enrichment Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## 1. Introduction

The enrichment analysis module performs metabolite set enrichment analysis (MSEA) for human and mammalian species based on several libraries containing ~6300 groups of metabolite sets. Users can upload either 1) a list of compounds, 2) a list of compounds with concentrations, or 3) a concentration table.

## 2. Enrichment Analysis Workflow

### 2.1 Over representation analysis

We will go over two analysis workflows, the first is when the input is a list to perform over representation analysis. The first step is to create a vector containing a list of compound names. The list will then be cross-referenced (*CrossReferencing*) against the MetaboAnalyst compound libraries (HMDB, PubChem, KEGG, etc.), and any compounds without a hit will have NA. This step may take long due to downloading of libraries if they do not already exist in your working directory.

```{r, eval=FALSE}
## When input is a list

# Create vector consisting of compounds for enrichment analysis 
tmp.vec <- c("Acetoacetic acid", "Beta-Alanine", "Creatine", "Dimethylglycine", "Fumaric acid", "Glycine", "Homocysteine", "L-Cysteine", "L-Isolucine", "L-Phenylalanine", "L-Serine", "L-Threonine", "L-Tyrosine", "L-Valine", "Phenylpyruvic acid", "Propionic acid", "Pyruvic acid", "Sarcosine")

# Create mSetObj
mSet<-InitDataObjects("conc", "msetora", FALSE)

#Set up mSetObj with the list of compounds
mSet<-Setup.MapData(mSet, tmp.vec);

# Cross reference list of compounds against libraries (hmdb, pubchem, chebi, kegg, metlin)
mSet<-CrossReferencing(mSet, "name");

# View name map to identify any compounds within the uploaded list that do not have any hits
mSet$name.map 

$hit.inx
 [1]   29   86  199  221 6070  236  252 1142   NA 1224 1161 1252 1200 1251  330  349  364  372

$hit.values
 [1] "Acetoacetic acid"   "Beta-Alanine"       "Creatine"           "Dimethylglycine"    "Fumaric acid"       "Glycine"           
 [7] "Homocysteine"       "L-Cysteine"         NA                   "L-Phenylalanine"    "L-Serine"           "L-Threonine"       
[13] "L-Tyrosine"         "L-Valine"           "Phenylpyruvic acid" "Propionic acid"     "Pyruvic acid"       "Sarcosine"         

$match.state
 [1] 1 1 1 1 1 1 1 1 0 1 1 1 1 1 1 1 1 1

# Create the mapping results table
mSet<-CreateMappingResultTable(mSet)

# Input the name of the compound without any matches 
mSet<-PerformDetailMatch(mSet, "L-Isolucine");

# Create list of candidates to replace the compound
mSet <- GetCandidateList(mSet);

# Identify the name of the compound to replace
mSet<-SetCandidate(mSet, "L-Isolucine", "L-Isoleucine");

# Set the metabolite filter
mSet<-SetMetabolomeFilter(mSet, F);

# Select metabolite set library, refer to 
mSet<-SetCurrentMsetLib(mSet, "pathway", 0);

# Calculate hypergeometric score, results table generated in your working directory
mSet<-CalculateHyperScore(mSet)

# Plot the ORA, bar-graph
mSet<-PlotORA(mSet, "ora_0_", "bar", "png", 72, width=NA)
```

### 2.2 Quantitative Enrichment Analysis

Below, we will go over a second analysis workflow to perform QEA, where data input is a table.

```{r, eval=FALSE}
# Create mSetObj
mSet<-InitDataObjects("conc", "msetqea", FALSE)

# Read in data table
mSet<-Read.TextData(mSet, "http://www.metaboanalyst.ca/MetaboAnalyst/resources/data/human_cachexia.csv", "rowu", "disc");

# Perform cross-referencing of compound names
mSet<-CrossReferencing(mSet, "name");

# Create mapping results table
mSet<-CreateMappingResultTable(mSet)

# Mandatory check of data 
mSet<-SanityCheckData(mSet);

# Replace missing values with minimum concentration levels
mSet<-ReplaceMin(mSet);

# Perform no normalization
mSet<-Normalization(mSet, "NULL", "NULL", "NULL", "PIF_178", ratio=FALSE, ratioNum=20)

# Plot normalization
mSet<-PlotNormSummary(mSet, "norm_0_", "png", 72, width=NA)

# Plot sample-wise normalization
mSet<-PlotSampleNormSummary(mSet, "snorm_0_", "png", 72, width=NA)

# Set the metabolome filter
mSet<-SetMetabolomeFilter(mSet, F);

# Set the metabolite set library to pathway
mSet<-SetCurrentMsetLib(mSet, "pathway", 0);

# Calculate the global test score
mSet<-CalculateGlobalTestScore(mSet)

# Plot the QEA
mSet<-PlotQEA.Overview(mSet, "qea_0_", "bar", "png", 72, width=NA)

```

## 3. Sweave Report 

To prepare the sweave report, please use the *CreatePDFReport* function. You must ensure that you have the nexessary Latex libraries to generate the report (i.e. pdflatex, LaTexiT). The object created *must* be named *mSet*, and specify the user name in quotation marks. 
